### 2016-09-14 paper.tex
Updating conflict resolution

\title[PGVD]%
      {Parallel Octree Construction on Collections of Objects}

@@

\begin{abstract}
We present a parallel octree algorithm that resolves between geometric objects. The octree has the property that no octree cell intersects more than one labeled object. Previous parallel algorithms either spawn kernels hierarchically, separate points only, or make no hard guarantees of object separation. Our algorithm runs in \red{complexity?} in the average case and has excellent results in practice. We demonstrate with results on 2D and 3D datasets.

@@

We present an algorithm to compute a GVD approximation on arbitrary datasets, including those with closely spaced objects.  The approach applies a distance transform over an octree representation of the objects.  Our octree, its associated data structure, and our distance transform are novel and optimized to GVD approximation. For the remainder of the paper, ``GVD'' will refer to the approximated Generalized Voronoi Diagram.

@@

\begin{enumerate}
\item Construct an octree on object vertices using Karras' algorithm \cite{karras2012maximizing}
\item Detect octree cells that intersect more than one object, which we call ``conflict cells'' (contribution)
\item Subdivide conflict cells to resolve objects (contribution)
\end{enumerate}

@@

\paragraph{Generalized Voronoi diagrams}
A theoretical framework for generalized Voronoi diagrams can be found in Boissonnat \etal \shortcite{boissonnat2006curved}. Ordinary Voronoi diagrams are well studied and efficient algorithms exist that compute them exactly \cite{de2008computational}, but exact algorithms for the generalized Voronoi diagram are limited to a small set of special cases \cite{lee1982medial,karavelas2004robust}. In an early work, Lavender \etal \shortcite{lavender1992voronoi} define and compute GVDs over a set of solid models using an octree.  Etzion and Rappoport \shortcite{etzion2002computing} represent the GVD bisector symbolically for lazy evaluation, but are limited to sites that are polyhedra.  Boada \etal \shortcite{boada2002voronoi,boada2008approximations} use an adaptive approach to GVD computation, but their algorithm is restricted to GVDs with connected regions and is inefficient for polyhedral objects with many facets.  Two other works are adaptive \cite{teichmann1997polygonal,vleugels1998approximating} but are computationally expensive and are restricted to convex sites.

@@

\paragraph{Distance fields and octrees}
The GVD is a subset of the locus of distance field critical points, a property that we take advantage of. In that light, the GVD could be a post-processing step to any method that computes a distance field.  Distance transforms compute a distance field, but most are uniformly gridded \cite{jones20063d} and are thus no more suitable than GVD algorithms that use the GPU.

% JME: note that in the Bastos work, ``reconstruction'' is reconstruction
% of the surface from the octree at rendering time.
Two seminal works adaptively compute the Adaptive Distance Field (ADF) on octree vertices.  Strain~\shortcite{strain1999fast} fully resolves the octree everywhere on the object surface, and Frisken \etal~\shortcite{frisken2000adaptively} resolve the octree fully only in areas of small local feature size.  Both approaches are designed to retain features of a single object rather than resolving between multiple objects, as is required for GVD computation.  Qu \etal \shortcite{qu2004feature} implement an energy-minimizing distance field algorithm that preserves features at the expense of efficiency.  Many recent works on fast octree construction using the GPU are limited to point sites \cite{bedorf2012sparse,karras2012maximizing,zhou2011data}. Most octree approaches that support surfaces \cite{baert2013out,crassin2009gigavoxels,laine2011efficient,lefebvre2007compressed} are designed for efficient rendering, and actual construction of the octree is implemented on the CPU.

Two works \cite{bastos2008gpu,park2010cuda} implement the ADF using GPU parallelism to compute the distance value at sample points, but building the octree itself is done sequentially.  Yin \etal~\shortcite{yin2011fast} compute the distance field entirely on the GPU using a bottom-up approach by initially subdividing into a complete octree, resulting in memory usage that is no better than using a uniform grid.  A method by Kim and Liu~\shortcite{kim2014exact} computes the octree and a BVH entirely on the GPU. However, octree construction is performed on barycenters of triangles, and so a leaf octree cell can have an arbitrary number of triangle intersections as long as it contains no more than one triangle's barycenter.  We have found no GPU octree construction method that can resolve between objects.

%-----------------------------------------------------------
% Build octree
%-----------------------------------------------------------
\section{Algorithm}
\label{sec:algorithm}
Our algorithm works in both 2D and 3D. Lacking a dimension-independent term, we use ``octree'' as a general term to refer to both quadtrees and octrees. We refer to object facets, which are lines in 2D and triangles in 3D.

@@

%\begin{adjustbox}{valign=t}
\begin{figure}
  \centering
  \subfloat[][]{
    \label{fig:conflict-resolution-x}
    \includegraphics[width=0.48\columnwidth]{conflict-resolution-x.pdf} }
  \subfloat[][]{
    \label{fig:conflict-resolution-even}
    \includegraphics[width=0.48\columnwidth]{conflict-resolution-even.pdf} } \\
  \subfloat[][]{
    \label{fig:conflict-resolution-all}
    \includegraphics[width=0.48\columnwidth]{conflict-resolution-all.pdf} }
  \subfloat[][]{
    \label{fig:conflict-resolution-octree}
    \includegraphics[width=0.48\columnwidth]{conflict-resolution-octree-samples.pdf} }
  \caption{
    \protect\subref{fig:conflict-resolution-x} A conflict cell with two lines from different objects.
    \protect\subref{fig:conflict-resolution-even} Fitting boxes such that any box intersecting both lines contains at least one sample (red dots).
    \protect\subref{fig:conflict-resolution-even} Fitting boxes such that any box intersecting both lines contains at least two samples. This ensures that an octree built from the samples using Karras' algorithm (panel \protect\subref{fig:conflict-resolution-octree}) will have no leaf cells that intersect both lines, ensuring that the new octree is locally free of conflict cells.
  }
  \label{fig:conflict-resolution}
\end{figure}
%    \end{adjustbox}

@@

To resolve conflict cells we consider pairs of lines of differing labels that intersect a cell $c$. Figure \ref{fig:conflict-resolution-x} shows two lines

\begin{align}
q(t) &= q = q_0 + tv \label{eqn:q} \\
r(f) &= r = r_0 + fw \label{eqn:r}
\end{align}
along with a line
\begin{align}
p(s) &= p = p_0 + su \label{eqn:p}
\end{align}
Our strategy will be to sample points $P$ on $p(u)$ (figure \ref{fig:conflict-resolution-octree}) such that an octree built on $V \cup P$ will completely ``separate'' $q$ and $r$, i.e., no descendent cell of $c$ will intersect both $q$ and $r$. We do this by ensuring that $P$ is sampled such that every box that intersects both $q$ and $r$ also intersects at least two points in $P$. Because Karras' algorithm guarantees that every leaf cell intersects at most one point, we know that no leaf cell will intersect $q$ and $r$ and thus no leaf cell will be a conflict cell.

We consider only cases where the slope of $p$ is in the range $0 \le m \le 1$. All other cases can be transformed to this case using rotation and reflection. We begin by fitting the smallest box centered on a point $p$ that intersects both $q$ and $r$.

\subsubsection{Case $w^y>0$}

@@

%
%. When $i$ is odd, the initial condition will be where $s$ is halfway along the first box (it could be anywhere in the interior of the first box, but we use halfway for convenience). So from \eqref{eqn:p},
%\begin{align}
%s_1 &= \frac{p^x-p_0^x}{2u^x} \\
%    &= \frac{a(0)}{2u^x}
%\end{align}
%\begin{align}
%k_1^{even} = k_1^{odd} &= \beta\left(1-\frac{\alpha}{\alpha-1}\right) \\
%k_2^{even} &= \frac{\beta}{\alpha-1} \\
%k_2^{odd} &= \frac{a(0)}{2u^x}+\frac{\beta}{\alpha-1}
%\end{align}
%
%It is tempting to formulate the recurrance relation using a single initial condition. This would be done by replacing equation \eqref{eqn:xi_init} with
%\begin{align}
%x_i &= x_{i-1} + \frac{a(s_{i-1})}{2}
%\end{align}
%This formulation can lead to conflict cells as shown in figure \ref{fig:conflict-resolution-half}.
%
%\begin{figure}
%  \centering
%  \includegraphics[width=0.48\columnwidth]{conflict-resolution-half.pdf}
%  \caption{Using a single recurrence solution we can no longer guarantee that every box that intersects both lines contains at least two samples. The box in red contains only one sample, so the octree may contain conflict cells.
%  }
%  \label{fig:conflict-resolution-half}
%\end{figure}

@@

\subsection{Build octree on vertices}

We first construct an octree on the vertices of the objects, which we call the ``vertex octree''. We use Karras' algorithm \cite{karras2012maximizing} which sorts the Morton codes of the vertices in parallel, then constructs the binary radix tree in parallel. With the binary radix tree, the octree can be constructed with a single parallel call. The strength of this algorithm lies in the fact that overall performance scales linearly with the number of cores, regardless of the distribution of points. That is, even if a large number of vertices are clustered in a small area, requiring deep octree subdivision, only a constant number of parallel calls need be made. Given enough parallel units, the Karras algorithm runs in $O(\log{N})$ time, where $N$ is the number of vertices.

@@

Our end goal is to construct an octree such that no octree cell intersects more than one object. Note that a cell is allowed to intersect more than one facet, but all facets must belong to the same object, or, in other words, all facets must share the same label. It is possible, but unlikely, that the vertex octree has this property. If so, then we are done. Otherwise, we must identify subdivide conflict cells.

@@

We identify conflict cells as shown in algorithm \ref{alg:find-conflict-cells}. In lines \ref{alg:octree_containing_begin}-\ref{alg:octree_containing_end}, for each internal octree cell $c$, we store all facets for which $c$ is the smallest containing cell. Since we are implementing this in a GPGPU environment, we don't have dynamic memory, so each facet must be processed twice. The first loop discovers how many facets are to be stored in each cell after which we allocate space for the facets. We use parallel prefix sums to determine the amount of space we need to allocate as well as the offsets for each internal cell. The second loop actually stores the facets.

The $\container{f}$ procedure finds the smallest octree cell that fully contains the facet $f$. A straightforward implementation of $\container{f}$ is to perform a standard octree search on the vertices of $f$ and take the smallest octree cell that contains all of them. (Note that the cell is always an internal node, since a post-condition of the Karras algorithm is that no leaf cell contains more than one vertex.) In our implementation however, we take advantage of our existing data structures. The octree cell that contains a vertex $v$ is uniquely determined by the D-tuple bits of its morton code. For example, if a 2D vertex has morton code $010010$, then the octree is traversed from the root to child $01$ to child $00$ to child $10$. To determine $\container{f}$, we find the longest common prefix (\lcp) of the vertices. Truncating the length of \lcp\,to a multiple of D, we find the smallest octree cell that contains all vertices of $f$. The complexity of $\container{f}$ is $O(\log{N})$ for both implementations. Thus, lines \ref{alg:octree_containing_begin}-\ref{alg:octree_containing_end} run in $O(\log{N})$ time.

Lines \ref{alg:octree_intersections_begin}-\ref{alg:octree_intersections_end} of the algorithm identify and store all facets that intersect with a given leaf cell $c$. Again, it is done in two steps for memory allocation purposes. Each leaf cell $c$ looks at its $O(\log{N})$ ancestors and tests all facets stored in those ancestors for intersection with $c$. Any intersecting facets get stored in $c$. These lines run in $O(F)$ time, where $F$ is the number of facets in all objects. Even though the loop is doubly-nested, each facet is stored in a unique internal node, so no more than $F$ facets will be visited in the loops. In practice, far fewer than $F$ facets will be checked for each leaf cell, because most datasets have facets that are completely contained in internal cells that are reasonably low in the tree.

@@

In Step 4, Stack is preallocated to size $M\cdot2^D$ where $M$ is the maximum octree depth and $D$ is the dimension. A conflict cell is a cell that intersects at least two different objects, or two lines of different labels.

The second procedure we use is $\directAncestors{c}$, which finds all ancestors of octree cell $c$.

@@

\algorithmspace
\begin{algorithm}
  \DontPrintSemicolon
  \LinesNumbered
  \KwIn{VertexOctree}
  \BlankLine
  \tcp{Store contained facets}
  \ForPar{facet $f$ in Objects}{
    \label{alg:octree_containing_begin}
    a := $\container{f}$\;
    a.numFacets := a.numFacets + 1\;
  }
  Allocate space for facets in internal cells\;
  \ForPar{facet $f$ in Objects}{
    a := $\container{f}$\;
    a.facets := a.facets $\cup$ f
  } \label{alg:octree_containing_end}
  \tcp{Store intersecting facets}
  \ForPar{leaf cell c in VertexOctree}{
    \label{alg:octree_intersections_begin}
    \ForEach{cell a in \directAncestors{c}}{
      \ForEach{facet $f$ in a.facets}{
        \If{$f$ intersects c}{
          c.numFacets := c.numFacets + 1\;
        }
      }
    }
  }
  Allocate space for facets in leaf cells\;
  \ForPar{leaf cell c in VertexOctree}{
    \ForEach{cell a in \directAncestors{c}}{
      \ForEach{facet $f$ in a.facets}{
        \If{$f$ intersects c}{
          c.facets := c.facets $\cup$ f\;
        }
      }
    }
  } \label{alg:octree_intersections_end}
\caption{FIND\_CONFLICT\_CELLS}
\label{alg:find-conflict-cells}
\end{algorithm}
\algorithmspace

\algorithmspace
\begin{algorithm}
  \DontPrintSemicolon
  \LinesNumbered
  \KwIn{Octree, conflict\_cells}
  \BlankLine
  \tcp{4. Octree refinement}
  \ForPar{leaf cell c in Octree}{
    c' := c\;
    \While{c' $\in$ conflict\_cells}{
      $(c'_0,c'_1,\dots,c'_{2^D-1})$ := subdivide c'\;
      push $(c'_0,c'_1,\dots,c'_{2^D-1})$ onto Stack\;
      c' := Stack.pop
    }
  }
\caption{REFINE\_OCTREE}
\label{alg:refine-octree}
\end{algorithm}
\algorithmspace

\begin{figure}
  \centering
  \subfloat[][]{
    \label{fig:octree-cartesian-1}
    \includegraphics[width=0.48\columnwidth]{octree-cartesian-1.pdf} }
  \subfloat[][]{
    \label{fig:octree-cartesian-2}
    \includegraphics[width=0.48\columnwidth]{octree-cartesian-2.pdf} } \\
  \subfloat[][]{
    \label{fig:octree-cartesian-3}
    \includegraphics[width=0.48\columnwidth]{octree-cartesian-3.pdf} }
  \subfloat[][]{
    \label{fig:octree-cartesian-5}
    \includegraphics[width=0.48\columnwidth]{octree-cartesian-5.pdf} } \\
  \subfloat[][]{
    \label{fig:octree-hierarchical-1}
    \includegraphics[width=0.48\columnwidth]{octree-hierarchical-1.pdf} }
  \subfloat[][]{
    \label{fig:octree-hierarchical-2}
    \includegraphics[width=0.48\columnwidth]{octree-hierarchical-2.pdf} }
  \caption{
    \protect\subref{fig:octree-cartesian-1} A red object and a blue object.
    \protect\subref{fig:octree-cartesian-2} The vertex octree, or octree built on the object vertices using Karras' algorithm.
    \protect\subref{fig:octree-cartesian-3}, \protect\subref{fig:octree-hierarchical-1} The vertex octree with conflict cells highlighted. Note the label of an octree cell in \protect\subref{fig:octree-cartesian-3} is the concatenation of labels from root R to the leaf cell in \protect\subref{fig:octree-hierarchical-1}. This value also corresponds to the highest order bits of the morton code of any point in the cell.
    \protect\subref{fig:octree-cartesian-5}, \protect\subref{fig:octree-hierarchical-2} The octree after resolution of conflict cells.
  }
  \label{fig:steps}
\end{figure}

@@

% # octree vertices
%  52K ./viewer3 -l 8 ~/data/gears/gear[1-3].obj
% 170K ./viewer3 -l 12 ~/data/knife/knife-holder.obj ~/data/knife/knife[2-4].obj
% 146K ./viewer2 -l 24 ../data2/filled_ut/*.dat
% 2.8M ./viewer3 --uniform-colors -a 1 -l 8 ~/data/neuron/improved/a*.obj
% 1.3M ./viewer3 -l 8 ~/data/rice-dwarf/decimated2/n1UF2a-*.obj

% memory (54 bytes per octree cell)
%  2.8Mb ./viewer3 -l 8 ~/data/gears/gear[1-3].obj
%  9.2Mb ./viewer3 -l 12 ~/data/knife/knife-holder.obj ~/data/knife/knife[2-4].obj
% 7.9Mb ./viewer2 -l 24 ../data2/filled_ut/*.dat
% 151.2Mb ./viewer3 --uniform-colors -a 1 -l 8 ~/data/neuron/improved/a*.obj
% 70.2Mb ./viewer3 -l 8 ~/data/rice-dwarf/decimated2/n1UF2a-*.obj

>>>>

\title[PGVD]%
      {Parallel Quadtree Construction on Collections of Objects}

@@

\begin{abstract}
We present a parallel quadtree algorithm that resolves between geometric objects. The quadtree has the property that no quadtree cell intersects more than one labeled object. Previous parallel algorithms either spawn kernels hierarchically, separate points only, or make no hard guarantees of object separation. Our algorithm runs in \red{complexity?} in the average case and has excellent results in practice. We demonstrate with results on 2D and 3D datasets.

@@

We present an algorithm to compute a GVD approximation on arbitrary datasets, including those with closely spaced objects.  The approach applies a distance transform over an quadtree representation of the objects.  Our quadtree, its associated data structure, and our distance transform are novel and optimized to GVD approximation. For the remainder of the paper, ``GVD'' will refer to the approximated Generalized Voronoi Diagram.

@@

\begin{enumerate}
\item Construct an quadtree on object vertices using Karras' algorithm \cite{karras2012maximizing}
\item Detect quadtree cells that intersect more than one object, which we call ``conflict cells'' (contribution)
\item Subdivide conflict cells to resolve objects (contribution)
\end{enumerate}

@@

\paragraph{Generalized Voronoi diagrams}
A theoretical framework for generalized Voronoi diagrams can be found in Boissonnat \etal \shortcite{boissonnat2006curved}. Ordinary Voronoi diagrams are well studied and efficient algorithms exist that compute them exactly \cite{de2008computational}, but exact algorithms for the generalized Voronoi diagram are limited to a small set of special cases \cite{lee1982medial,karavelas2004robust}. In an early work, Lavender \etal \shortcite{lavender1992voronoi} define and compute GVDs over a set of solid models using an quadtree.  Etzion and Rappoport \shortcite{etzion2002computing} represent the GVD bisector symbolically for lazy evaluation, but are limited to sites that are polyhedra.  Boada \etal \shortcite{boada2002voronoi,boada2008approximations} use an adaptive approach to GVD computation, but their algorithm is restricted to GVDs with connected regions and is inefficient for polyhedral objects with many facets.  Two other works are adaptive \cite{teichmann1997polygonal,vleugels1998approximating} but are computationally expensive and are restricted to convex sites.

@@

\paragraph{Distance fields and quadtrees}
The GVD is a subset of the locus of distance field critical points, a property that we take advantage of. In that light, the GVD could be a post-processing step to any method that computes a distance field.  Distance transforms compute a distance field, but most are uniformly gridded \cite{jones20063d} and are thus no more suitable than GVD algorithms that use the GPU.

% JME: note that in the Bastos work, ``reconstruction'' is reconstruction
% of the surface from the quadtree at rendering time.
Two seminal works adaptively compute the Adaptive Distance Field (ADF) on quadtree vertices.  Strain~\shortcite{strain1999fast} fully resolves the quadtree everywhere on the object surface, and Frisken \etal~\shortcite{frisken2000adaptively} resolve the quadtree fully only in areas of small local feature size.  Both approaches are designed to retain features of a single object rather than resolving between multiple objects, as is required for GVD computation.  Qu \etal \shortcite{qu2004feature} implement an energy-minimizing distance field algorithm that preserves features at the expense of efficiency.  Many recent works on fast quadtree construction using the GPU are limited to point sites \cite{bedorf2012sparse,karras2012maximizing,zhou2011data}. Most quadtree approaches that support surfaces \cite{baert2013out,crassin2009gigavoxels,laine2011efficient,lefebvre2007compressed} are designed for efficient rendering, and actual construction of the quadtree is implemented on the CPU.

Two works \cite{bastos2008gpu,park2010cuda} implement the ADF using GPU parallelism to compute the distance value at sample points, but building the quadtree itself is done sequentially.  Yin \etal~\shortcite{yin2011fast} compute the distance field entirely on the GPU using a bottom-up approach by initially subdividing into a complete quadtree, resulting in memory usage that is no better than using a uniform grid.  A method by Kim and Liu~\shortcite{kim2014exact} computes the quadtree and a BVH entirely on the GPU. However, quadtree construction is performed on barycenters of triangles, and so a leaf quadtree cell can have an arbitrary number of triangle intersections as long as it contains no more than one triangle's barycenter.  We have found no GPU quadtree construction method that can resolve between objects.

%-----------------------------------------------------------
% Build quadtree
%-----------------------------------------------------------
\section{Algorithm}
\label{sec:algorithm}

@@

%\begin{adjustbox}{valign=t}
\begin{figure}
  \centering
  \subfloat[][]{
    \label{fig:conflict-resolution-x}
    \includegraphics[width=0.48\columnwidth]{conflict-resolution-x.pdf} }
  \subfloat[][]{
    \label{fig:conflict-resolution-even}
    \includegraphics[width=0.48\columnwidth]{conflict-resolution-even.pdf} } \\
  \subfloat[][]{
    \label{fig:conflict-resolution-all}
    \includegraphics[width=0.48\columnwidth]{conflict-resolution-all.pdf} }
  \subfloat[][]{
    \label{fig:conflict-resolution-octree}
    \includegraphics[width=0.48\columnwidth]{conflict-resolution-octree.pdf} }
  \caption{
    \protect\subref{fig:conflict-resolution-x} A conflict cell with two lines from different objects.
    \protect\subref{fig:conflict-resolution-even} Fitting boxes such that any box intersecting both lines contains at least one sample (red dots).
    \protect\subref{fig:conflict-resolution-even} Fitting boxes such that any box intersecting both lines contains at least two samples. This ensures that an quadtree built from the samples using Karras' algorithm (panel \protect\subref{fig:conflict-resolution-octree}) will have no leaf cells that intersect both lines, ensuring that the new quadtree is locally free of conflict cells.
  }
  \label{fig:conflict-resolution}
\end{figure}
%    \end{adjustbox}

@@

A conflict cell is a quadtree cell that intersects at least two different objects. To resolve a conflict cell $c$, we consider pairs of lines of differing labels that intersect $c$. Figure \ref{fig:conflict-resolution-x} shows two lines

\begin{align}
q(t) &= q = q_0 + tv \label{eqn:q} \\
r(f) &= r = r_0 + fw \label{eqn:r}
\end{align}
along with a line
\begin{align}
p(s) &= p = p_0 + su \label{eqn:p}
\end{align}
that bisects $q$ and $r$. Our strategy will be to sample points $P$ on $p(u)$ (figure \ref{fig:conflict-resolution-octree}) such that an quadtree built on $V \cup P$ will completely ``separate'' $q$ and $r$, i.e., no descendent cell of $c$ will intersect both $q$ and $r$. We do this by ensuring that $P$ is sampled such that every box that intersects both $q$ and $r$ also intersects at least two points in $P$. Because Karras' algorithm guarantees that every leaf cell intersects at most one point, we know that no leaf cell will intersect $q$ and $r$ and thus no leaf cell will be a conflict cell.

We consider only cases where the slope of $p$ is in the range $0 \le m \le 1$. All other cases can be transformed to this case using rotation and reflection. We begin by fitting the smallest box centered on a point $p$ that intersects both $q$ and $r$. We break the problem into two cases: the \textit{opposite} case is where $w^y > 0$, so each box intersects $q$ and $r$ at its top-left and bottom-right corners, respectively. The \textit{adjacent} case  is where $w^y < 0$, so the line intersections are adjacent -- the top-left and bottom-left corners of the box. Figure \ref{fig:conflict-resolution-octree} shows the \textit{opposite} case.

\subsubsection{\textit{Opposite} case}

@@

%
%. When $i$ is odd, the initial condition will be where $s$ is halfway along the first box (it could be anywhere in the interior of the first box, but we use halfway for convenience). So from \eqref{eqn:p},
%\begin{align}
%s_1 &= \frac{p^x-p_0^x}{2u^x} \\
%    &= \frac{a(0)}{2u^x}
%\end{align}
%\begin{align}
%k_1^{even} = k_1^{odd} &= \beta\left(1-\frac{\alpha}{\alpha-1}\right) \\
%k_2^{even} &= \frac{\beta}{\alpha-1} \\
%k_2^{odd} &= \frac{a(0)}{2u^x}+\frac{\beta}{\alpha-1}
%\end{align}
%
%It is tempting to formulate the recurrance relation using a single initial condition. This would be done by replacing equation \eqref{eqn:xi_init} with
%\begin{align}
%x_i &= x_{i-1} + \frac{a(s_{i-1})}{2}
%\end{align}
%This formulation can lead to conflict cells as shown in figure \ref{fig:conflict-resolution-half}.
%
%\begin{figure}
%  \centering
%  \includegraphics[width=0.48\columnwidth]{conflict-resolution-half.pdf}
%  \caption{Using a single recurrence solution we can no longer guarantee that every box that intersects both lines contains at least two samples. The box in red contains only one sample, so the quadtree may contain conflict cells.
%  }
%  \label{fig:conflict-resolution-half}
%\end{figure}

@@

\subsection{Build quadtree on vertices}

We first construct an quadtree on the vertices of the objects, which we call the ``vertex quadtree''. We use Karras' algorithm \cite{karras2012maximizing} which sorts the Morton codes of the vertices in parallel, then constructs the binary radix tree in parallel. With the binary radix tree, the quadtree can be constructed with a single parallel call. The strength of this algorithm lies in the fact that overall performance scales linearly with the number of cores, regardless of the distribution of points. That is, even if a large number of vertices are clustered in a small area, requiring deep quadtree subdivision, only a constant number of parallel calls need be made. Given enough parallel units, the Karras algorithm runs in $O(\log{N})$ time, where $N$ is the number of vertices.

@@

Our end goal is to construct an quadtree such that no quadtree cell intersects more than one object. Note that a cell is allowed to intersect more than one facet, but all facets must belong to the same object, or, in other words, all facets must share the same label. It is possible, but unlikely, that the vertex quadtree has this property. If so, then we are done. Otherwise, we must identify subdivide conflict cells.

@@

We identify conflict cells as shown in algorithm \ref{alg:find-conflict-cells}. In lines \ref{alg:quadtree_containing_begin}-\ref{alg:quadtree_containing_end}, for each internal quadtree cell $c$, we store all facets for which $c$ is the smallest containing cell. Since we are implementing this in a GPGPU environment, we don't have dynamic memory, so each facet must be processed twice. The first loop discovers how many facets are to be stored in each cell after which we allocate space for the facets. We use parallel prefix sums to determine the amount of space we need to allocate as well as the offsets for each internal cell. The second loop actually stores the facets.

The $\container{f}$ procedure finds the smallest quadtree cell that fully contains the facet $f$. A straightforward implementation of $\container{f}$ is to perform a standard quadtree search on the vertices of $f$ and take the smallest quadtree cell that contains all of them. (Note that the cell is always an internal node, since a post-condition of the Karras algorithm is that no leaf cell contains more than one vertex.) In our implementation however, we take advantage of our existing data structures. The quadtree cell that contains a vertex $v$ is uniquely determined by the D-tuple bits of its morton code. For example, if a 2D vertex has morton code $010010$, then the quadtree is traversed from the root to child $01$ to child $00$ to child $10$. To determine $\container{f}$, we find the longest common prefix (\lcp) of the vertices. Truncating the length of \lcp\,to a multiple of D, we find the smallest quadtree cell that contains all vertices of $f$. The complexity of $\container{f}$ is $O(\log{N})$ for both implementations. Thus, lines \ref{alg:quadtree_containing_begin}-\ref{alg:quadtree_containing_end} run in $O(\log{N})$ time.

Lines \ref{alg:quadtree_intersections_begin}-\ref{alg:quadtree_intersections_end} of the algorithm identify and store all facets that intersect with a given leaf cell $c$. Again, it is done in two steps for memory allocation purposes. Each leaf cell $c$ looks at its $O(\log{N})$ ancestors and tests all facets stored in those ancestors for intersection with $c$. Any intersecting facets get stored in $c$. These lines run in $O(F)$ time, where $F$ is the number of facets in all objects. Even though the loop is doubly-nested, each facet is stored in a unique internal node, so no more than $F$ facets will be visited in the loops. In practice, far fewer than $F$ facets will be checked for each leaf cell, because most datasets have facets that are completely contained in internal cells that are reasonably low in the tree.

@@

In Step 4, Stack is preallocated to size $M\cdot2^D$ where $M$ is the maximum quadtree depth and $D$ is the dimension. A conflict cell is a cell that intersects at least two different objects, or two lines of different labels.

The second procedure we use is $\directAncestors{c}$, which finds all ancestors of quadtree cell $c$.

@@

\algorithmspace
\begin{algorithm}
  \DontPrintSemicolon
  \LinesNumbered
  \KwIn{VertexQuadtree}
  \BlankLine
  \tcp{Store contained facets}
  \ForPar{facet $f$ in Objects}{
    \label{alg:quadtree_containing_begin}
    a := $\container{f}$\;
    a.numFacets := a.numFacets + 1\;
  }
  Allocate space for facets in internal cells\;
  \ForPar{facet $f$ in Objects}{
    a := $\container{f}$\;
    a.facets := a.facets $\cup$ f
  } \label{alg:quadtree_containing_end}
  \tcp{Store intersecting facets}
  \ForPar{leaf cell c in VertexQuadtree}{
    \label{alg:quadtree_intersections_begin}
    \ForEach{cell a in \directAncestors{c}}{
      \ForEach{facet $f$ in a.facets}{
        \If{$f$ intersects c}{
          c.numFacets := c.numFacets + 1\;
        }
      }
    }
  }
  Allocate space for facets in leaf cells\;
  \ForPar{leaf cell c in VertexQuadtree}{
    \ForEach{cell a in \directAncestors{c}}{
      \ForEach{facet $f$ in a.facets}{
        \If{$f$ intersects c}{
          c.facets := c.facets $\cup$ f\;
        }
      }
    }
  } \label{alg:quadtree_intersections_end}
\caption{FIND\_CONFLICT\_CELLS}
\label{alg:find-conflict-cells}
\end{algorithm}
\algorithmspace

\algorithmspace
\begin{algorithm}
  \DontPrintSemicolon
  \LinesNumbered
  \KwIn{Quadtree, conflict\_cells}
  \BlankLine
  \tcp{4. Quadtree refinement}
  \ForPar{leaf cell c in Quadtree}{
    c' := c\;
    \While{c' $\in$ conflict\_cells}{
      $(c'_0,c'_1,\dots,c'_{2^D-1})$ := subdivide c'\;
      push $(c'_0,c'_1,\dots,c'_{2^D-1})$ onto Stack\;
      c' := Stack.pop
    }
  }
\caption{REFINE\_QUADTREE}
\label{alg:refine-quadtree}
\end{algorithm}
\algorithmspace

\begin{figure}
  \centering
  \subfloat[][]{
    \label{fig:octree-cartesian-1}
    \includegraphics[width=0.48\columnwidth]{octree-cartesian-1.pdf} }
  \subfloat[][]{
    \label{fig:octree-cartesian-2}
    \includegraphics[width=0.48\columnwidth]{octree-cartesian-2.pdf} } \\
  \subfloat[][]{
    \label{fig:octree-cartesian-3}
    \includegraphics[width=0.48\columnwidth]{octree-cartesian-3.pdf} }
  \subfloat[][]{
    \label{fig:octree-cartesian-5}
    \includegraphics[width=0.48\columnwidth]{octree-cartesian-5.pdf} } \\
  \subfloat[][]{
    \label{fig:octree-hierarchical-1}
    \includegraphics[width=0.48\columnwidth]{octree-hierarchical-1.pdf} }
  \subfloat[][]{
    \label{fig:octree-hierarchical-2}
    \includegraphics[width=0.48\columnwidth]{octree-hierarchical-2.pdf} }
  \caption{
    \protect\subref{fig:octree-cartesian-1} A red object and a blue object.
    \protect\subref{fig:octree-cartesian-2} The vertex quadtree, or quadtree built on the object vertices using Karras' algorithm.
    \protect\subref{fig:octree-cartesian-3}, \protect\subref{fig:octree-hierarchical-1} The vertex quadtree with conflict cells highlighted. Note the label of an quadtree cell in \protect\subref{fig:octree-cartesian-3} is the concatenation of labels from root R to the leaf cell in \protect\subref{fig:octree-hierarchical-1}. This value also corresponds to the highest order bits of the morton code of any point in the cell.
    \protect\subref{fig:octree-cartesian-5}, \protect\subref{fig:octree-hierarchical-2} The quadtree after resolution of conflict cells.
  }
  \label{fig:steps}
\end{figure}

@@

% # quadtree vertices
%  52K ./viewer3 -l 8 ~/data/gears/gear[1-3].obj
% 170K ./viewer3 -l 12 ~/data/knife/knife-holder.obj ~/data/knife/knife[2-4].obj
% 146K ./viewer2 -l 24 ../data2/filled_ut/*.dat
% 2.8M ./viewer3 --uniform-colors -a 1 -l 8 ~/data/neuron/improved/a*.obj
% 1.3M ./viewer3 -l 8 ~/data/rice-dwarf/decimated2/n1UF2a-*.obj

% memory (54 bytes per quadtree cell)
%  2.8Mb ./viewer3 -l 8 ~/data/gears/gear[1-3].obj
%  9.2Mb ./viewer3 -l 12 ~/data/knife/knife-holder.obj ~/data/knife/knife[2-4].obj
% 7.9Mb ./viewer2 -l 24 ../data2/filled_ut/*.dat
% 151.2Mb ./viewer3 --uniform-colors -a 1 -l 8 ~/data/neuron/improved/a*.obj
% 70.2Mb ./viewer3 -l 8 ~/data/rice-dwarf/decimated2/n1UF2a-*.obj
